---
title: "Proteomic Landscape of NPM1 mutated AML - module 1 - data carpentry"
output: github_output
---

# Introduction

Given it's the largest molecular subgroup, molecular elucidation of NPM1^mut^ AML from the large proteomics dataset is warranted. To this end, I'll aim to give a detailed exploration of the proteomic landscape of NPM1^mut^ AML i) in comparison to the NPM^wt^ AML as well as ii) within the NPM1^mut^ group of leukemias.

Pathological background:

# Prerequisits

```{r}
library(tidyverse)
```

# Clinical Data and describtion of cohorts

Two different cohorts, AML Mito/FFM (*n* = 177) and SAL (*n* = 100) of untreated, newly diagnosed AML patients. Proteomics was done by means of label-free quantification with data indepentend acquisition (DIA) in two different batches (all AML Mito samples, all SAL samples).

```{r}
phen_data_lfq <- readxl::read_excel("~/Forschung/AG Oellerich/Datensätze/FFM_Discovery_Cohort/FFM_AML_Discovery_Cohort.xlsx")
phen_data_sal <- readxl::read_excel("~/Forschung/AG Oellerich/Datensätze/SALII_Cohort/KlinischeDaten_Kohorte_FFM_20220422.xlsx")
```

# Datasets

We'll work with the main LFQ dataset (n = ~ 177 pts) as well as with the SALII-revision dataset (n = 100 pts).

Pre-processing for the lfq was done as described in <https://github.com/jenssle/AG_Oellerich_ProtGen/blob/main/Proteomic_data_processing/20220306_lfq_proteome_normalization_cleanup.md>

and data merge and imputation in

<https://github.com/jenssle/AG_Oellerich_ProtGen/blob/main/Proteomic_data_processing/20220413_lfq_proteome_merge/20221413_lfq_proteome_preprocess_merge_impute.Rmd>

Briefly, samples with <30% coverage and proteins with >50% missingness were removed as were samples with obviously skewed distribution via visual inspection (4 AML, 1 CD34 sample removed). 

Notably, the data set also includes proteomics of sorted CD34 HSCs from 12 healthy donors (initially 13, 1 failed QC)

There're several protein/gene IDs to use as identifiers.

PG.Genes and PG.UniProtIds have approx. 350 NAs, the majority of which are NA in both PG.Genes and PG.UniProtIds. However, they all have a PG.ProteinAccession which can be used and which, in case of proteins with a PG.UniProtId, matches this. Furthermore, in cases where PG.UniProtIds is disambiguous (i.e. two ids separated by ;), both are disambiguous.

Imputation was done on the cluster using the DreamAI framework

```{r}
prot_merge_complete <- readRDS("~/Forschung/AG Oellerich/Datensätze/Proteomics Merge/20220608_prot_ffm_sal_cd34_merge_imputed.RDS")
```

The complete data set includes also the 12 healthy CD34 cells. 

## Intensity comparison batches

```{r}
prot_merge_complete %>% 
  as_tibble(rownames = "uniprot") %>% 
  pivot_longer(cols = -uniprot, names_to = "ID", values_to = "vsn_intensity") %>% 
  mutate(batch = case_when(
    stringr::str_starts(ID, "F") ~ "FFM",
    stringr::str_starts(ID, "AMLR") ~ "SAL",
    stringr::str_detect(ID, "CD34cells") ~ "CD34"
  )) %>% 
  ggplot(aes(x=vsn_intensity, col = batch)) +
  geom_density(size  = 1.5) +
  cowplot::theme_cowplot() +
  labs(title = "Sampel intensity comparison", caption = Sys.Date(), x = "VSN intensity", y = "Density")

ggsave(filename = "Sample Intensity comparison.jpg", device = "jpg")
```



