---
title: "20220517_proteomic_landscape_npm1_surface_proteome"
output: html_notebook
---
# Building a surface subproteome

In order to identify surface proteins of interest (which have high function in biomedical resarch since they might serve as antigens for targeted therapies or enhance diagnostics) we'll make use of the human Surfaceome by Bausch-Fluck et al., PlosOne 2015 (doi <https://doi.org/10.1371/journal.pone.0121314>). This resource contains a downloadable list of validated human surfaceome proteins (<http://wlab.ethz.ch/cspa/#downloads>).

The list includes the uniprotid, the up entry and protein name, furthermore whether the protein is identified to be a "CD" and the CSPA category confidence

regarding the confidence: Proteins with UniProt keywords Cell junction, Cell membrane, Secreted, (Signal peptide AND GPI-anchor) or (Signal peptide AND Transmembrane) were regarded as high-confidence surfaceome proteins (1039 human and 881 mouse proteins, Table A and B in S2 File). 247 human and 242 mouse proteins from the remaining pool have predicted transmembrane domains (but none of the above keywords assigned) and are therefore regarded as putative surfaceome proteins. The functional classification of these proteins and the empirical data of this study provide a basis to refine the annotation of the surfaceome. A remaining set of 206 human and 173 mouse proteins were classified as unspecific. These unspecific proteins were co-purified mainly due to their high abundance in the biological samples. In contrast to affinity enriched N-glycosites, chemical deamidation of asparagines is known to occur in solution. Among these co-purified peptides are nevertheless proteins linked to extracellular matrix (as procollagen transferases) and cytoskeletal protein (as tubulin), which could very well be associated with or present at the cell surface. Overall, the composite surfaceome contains 240/197 CD proteins, 346/315 receptors; of which 69/67 are G-protein coupled receptors and 39/36 receptor tyrosine kinases, and 184/207 transporters (human/mouse).

we'll only use high confidence surface proteins, this retains 234 proteins which are idenified in the data set.

```{r}
surfaceome <- read_delim("~/Forschung/AG Oellerich/Datensätze/Proteomics Surfacome/Surface Proteins.txt",delim =  "\t")

filter(surfaceome, `CSPA category` == "1 - high confidence") -> surfaceome_high_confidence


rownames(prot_merge) %in% surfaceome_high_confidence$`ID link` -> prot_merge_surface_vector
```

we'll store everyhing in a list object for classification on the cluster

```{r}
list(
  prot <- prot_merge,
  class_vector <- vector_CoI,
  surface_proteome <- prot_merge_surface_vector
) -> pred_prot

saveRDS(pred_prot, "pred_prot.RDS")
```



# LM DE surface proteins

surfaceome proteome is retrieved as described on the bottom 

```{r}
npm1_fit %>% 
  topTable(adjust.method = "BH", p.value = 0.01, number = Inf) %>% 
  rownames_to_column("PG.UniProtIds") %>% 
  left_join(
    npm1_fit_t_scores_biomart, by=c("PG.UniProtIds" = "uniprotswissprot")
  ) %>% 
  filter(PG.UniProtIds %in% surfaceome_high_confidence$`ID link`) %>% 
  mutate(in_consensus = factor(hgnc_symbol %in% c(consensus_signature_up_in_npm1, consensus_signature_down_in_npm1))) %>% 
  ggplot(aes(x=logFC, y= -log(adj.P.Val), label = hgnc_symbol, col = in_consensus)) +
  geom_point(alpha = .7) +
  cowplot::theme_cowplot() +
  ggrepel::geom_text_repel(max.overlaps = 50) +
  scale_color_manual(name = "Consensus", values = c("black", "#e85d04")) +
  scale_alpha_manual(values=c(1,1,0.3)) +
  theme(legend.position = "none") -> p_surface
```



# XGB Surfaceome Prediction

Florian Buettner trained an xgboost classifier for predicting NPM1 mutation status based on the surfaceome subproteome. An example setup is found below. 

Specifically, an 10fold nested cv was used to train a classifier and extract features of importance that appear in at least 9 of the 10 outer folds.  

```{r}
xgb_pred_npm1_surfaceome <- read_csv("xgb_npm1_pred_surfaceome.csv")

xgb_pred_npm1_surfaceome %>% 
  rename("uniprotid" = "...1") %>% 
  mutate(hgnc_symbol = uniprotswissprot_to_hgnc_symbol[uniprotid]) -> xgb_pred_npm1_surfaceome
  
```


We can use these proteins to visualize a heatmap


```{r}
prot_merge[xgb_pred_npm1_surfaceome$uniprotid, ] %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  ComplexHeatmap::Heatmap(top_annotation = NPM1_anno)
```














simple stepwise logit with DE surface proteins (refined approach with xgboost)

```{r}
npm1_fit %>% 
  topTable(adjust.method = "BH", p.value = 0.01, number = Inf) %>% 
  rownames_to_column("PG.UniProtIds") %>% 
  left_join(
    npm1_fit_t_scores_biomart, by=c("PG.UniProtIds" = "uniprotswissprot")
  ) %>% 
  filter(PG.UniProtIds %in% surfaceome_high_confidence$`ID link`) %>% 
  dplyr::select(PG.UniProtIds, hgnc_symbol) -> npm1_de_surfaceome


prot_merge[npm1_de_surfaceome$PG.UniProtIds, ] %>% 
  t() %>% 
  as_tibble(rownames = "ID") %>% 
  left_join(dplyr::select(phen_data, ID, NPM1)) %>% 
  mutate(NPM1 = factor(NPM1, levels=c(0,1), labels=c("wt", "mut"))) %>% 
  dplyr::select(-ID) -> npm1_de_surfaceome


MASS::stepAIC(glm(data = npm1_de_surfaceome, family = "binomial", formula = NPM1 ~ ., control = list(maxit = 1000)), trace = F) -> npm1_de_stepwise_logit

coef(npm1_de_stepwise_logit) %>% 
  as_tibble(rownames = "uniprotids") %>% 
  slice(-1) -> npm1_de_stepwise_logit 

npm1_de_stepwise_logit %>% 
  mutate(uniprotids = uniprotswissprot_to_hgnc_symbol[uniprotids])
  
  

```

using a simple 10fold outer cv xgboost to classify NPM1^mut^ vs. NPM1^wt^ on the surface proteome

```{r}
library(tidymodels)
library(xgboost)


prot_merge[npm1_de_surfaceome$PG.UniProtIds, ] %>% 
  t() %>% 
  as_tibble(rownames = "ID") %>% 
  left_join(
    dplyr::select(phen_data, ID, NPM1), by = c("ID")
  ) %>% 
  dplyr::select(-ID) %>% 
  mutate(NPM1 = factor(NPM1, levels=c(0,1), labels=c("wt", "mut"))) %>% 
  filter(!is.na(NPM1)) %>% 
  initial_split(strata = NPM1) -> prot_split

# create a 60/40 training/test split
prot_training <- training(prot_split)
prot_testing <- testing(prot_split)

recipe(NPM1 ~ ., data = prot_training) %>% 
  step_corr(all_predictors()) %>% 
  step_zv(all_predictors())-> xgb_recipe

# set model specs, tune everything

xgb_spec <- 
  boost_tree(
    mode = "classification",
    engine = "xgboost",
    trees = 1000,
    tree_depth = tune(),
    min_n = tune(),
    loss_reduction = tune(),
    sample_size = tune(),
    mtry = tune(),
    learn_rate = tune()
  )

# create a hypercube for the grid search
xgb_grid <-
  grid_latin_hypercube(
    tree_depth(), 
    min_n(), 
    loss_reduction(), 
    sample_size = sample_prop(), 
    finalize(mtry(), prot_training), 
    learn_rate(), 
    size = 100
  )

# create workflow
xgb_wf <- 
  workflow() %>% 
  add_recipe(xgb_recipe) %>% 
  add_model(xgb_spec) 

prot_fold <- vfold_cv(prot_training, strata = NPM1)

doParallel::registerDoParallel(cores  = parallel::detectCores() - 2)

xgb_res <- 
  tune_grid(
    xgb_wf,
    resamples = prot_fold,
    grid = xgb_grid,
    control = control_grid(save_pred = T)
  )

collect_metrics(xgb_res)
show_best(xgb_res, "roc_auc")

select_best(xgb_res, "roc_auc")

finalize_workflow(xgb_wf, select_best(xgb_res, "roc_auc")) -> xgb_final
```

Extract feature importance

```{r}
xgb_final %>% 
  fit(data = prot_training) %>% 
  pull_workflow_fit() %>% 
  vip::vip() -> xgb_vi

xgb_vi$data %>% 
  mutate(Variable = uniprotswissprot_to_hgnc_symbol[Variable]) %>% 
  ggplot(aes(x=forcats::fct_reorder(Variable, Importance), y = Importance)) +
  coord_flip() +
  geom_col() +
  cowplot::theme_cowplot() +
  labs(x="")
```

Fit the testing data set

```{r}
final_res <- last_fit(xgb_final, prot_split)
collect_metrics(final_res)
collect_predictions(final_res) %>% roc_curve(.pred_wt, truth = NPM1) %>% 
  autoplot()
```

